---
title: "Comparison to other software"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison to other software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## R code and results

```{r setup}
library(eventglm)

colon.cifit <- cumincglm(Surv(time, status) ~ rx, time = 2500, data = colon)
summary(colon.cifit)
confint(colon.cifit)
```


## Stata code and results

This uses the `st0202_1` package, available from here: https://www.stata-journal.com/article.html?article=st0202_1


```{stata, include = TRUE, eval = FALSE}
. import delimited "colon.csv", clear
(18 vars, 929 obs)

. 
. stset time, failure(status==1)

     failure event:  status == 1
obs. time interval:  (0, time]
 exit on or before:  failure

------------------------------------------------------------------------------
        929  total observations
          0  exclusions
------------------------------------------------------------------------------
        929  observations remaining, representing
        452  failures in single-record/single-failure data
  1,551,389  total analysis time at risk and under observation
                                                at risk from t =         0
                                     earliest observed entry t =         0
                                          last observed exit t =     3,329

. 
. // requires st0202_1 install (search stpci)
. stpci, at(2500)
Pseudo-observations for the cumulative incidence function.
Competing risks: (none).
Computing pseudo-observations (progress dots indicate percent completed).
----+--- 1 ---+--- 2 ---+--- 3 ---+--- 4 ---+--- 5 
..................................................    50
..................................................   100
Generated variable: pseudo.

. tabulate rx, gen(rxdum)

         rx |      Freq.     Percent        Cum.
------------+-----------------------------------
        Lev |        310       33.37       33.37
    Lev+5FU |        304       32.72       66.09
        Obs |        315       33.91      100.00
------------+-----------------------------------
      Total |        929      100.00

. glm pseudo rxdum1 rxdum2, vce(robust)

Iteration 0:   log pseudolikelihood =  -708.7556  

Generalized linear models                         Number of obs   =        929
Optimization     : ML                             Residual df     =        926
                                                  Scale parameter =    .270145
Deviance         =   250.154308                   (1/df) Deviance =    .270145
Pearson          =   250.154308                   (1/df) Pearson  =    .270145

Variance function: V(u) = 1                       [Gaussian]
Link function    : g(u) = u                       [Identity]

                                                  AIC             =   1.532305
Log pseudolikelihood = -708.7556003               BIC             =   -6078.23

------------------------------------------------------------------------------
             |               Robust
      pseudo |      Coef.   Std. Err.      z    P>|z|     [95% Conf. Interval]
-------------+----------------------------------------------------------------
      rxdum1 |   -.029075   .0416186    -0.70   0.485    -.1106459    .0524959
      rxdum2 |  -.1317578   .0417443    -3.16   0.002    -.2135752   -.0499404
       _cons |   .5434514     .02938    18.50   0.000     .4858676    .6010352
------------------------------------------------------------------------------
```

## SAS code and results

Assuming you have loaded the colon data in your workspace.

```{sas, eval = FALSE}
proc lifetest data=colon noprint plots=none timelist=2500 reduceout outsurv=sall;
time time*status(0);
run;

data sall;
set sall;
theta = survival;
keep theta;
run;

data sout;
set colon;
keep id;
run;

%macro pseudosurv;

%do ip=1 %to 929;
data coloni;
set colon;
where id ^= &ip;
run;

proc lifetest data=coloni noprint plots=none timelist=2500 reduceout outsurv=salli;
time time*status(0);
run;

data salli;
set salli;
thetamini = survival;
id = &ip;
keep id thetamini;
run;

data souti;
merge salli sall;
run;

data sout; 
merge sout souti;
by id;
run;
%end;
%mend pseudosurv;

%pseudosurv;


data sout2;
set sout;
pseudoci = 1 - (929 * theta - (929 - 1) * thetamini);
run;

data colon2;
merge colon sout2;
by id;
if rx='Lev' then rxlev = 1;
else rxlev = 0;
if rx='Lev+5FU' then rxlevplus = 1;
else rxlevplus = 0;
run;

proc reg data = colon2;
model pseudoci = rxlev rxlevplus / white;
run;

``` 



<div align="center">
<table class="table" cellspacing="0" cellpadding="5" rules="all" frame="box" bordercolor="#C1C1C1" summary="Procedure Reg: Parameter Estimates">
<colgroup>
<col>
<col>
</colgroup>
<colgroup>
<col>
<col>
<col>
<col>
<col>
<col>
<col>
</colgroup>
<thead>
<tr>
<th class="c b header" colspan="9" scope="colgroup">Parameter Estimates</th>
</tr>
<tr>
<th class="l b header" rowspan="2" scope="col">Variable</th>
<th class="r b header" rowspan="2" scope="col">DF</th>
<th class="r b header" rowspan="2" scope="col">Parameter<br/>Estimate</th>
<th class="r b header" rowspan="2" scope="col">Standard<br/>Error</th>
<th class="r b header" rowspan="2" scope="col">t&nbsp;Value</th>
<th class="r b header" rowspan="2" scope="col">Pr&nbsp;&gt;&nbsp;|t|</th>
<th class="c b header" colspan="3" scope="colgroup">Heteroscedasticity Consistent</th>
</tr>
<tr>
<th class="r b header" scope="col">Standard<br/>Error</th>
<th class="r b header" scope="col">t&nbsp;Value</th>
<th class="r b header" scope="col">Pr&nbsp;&gt;&nbsp;|t|</th>
</tr>
</thead>
<tbody>
<tr>
<th class="l rowheader" scope="row">Intercept</th>
<th class="r data">1</th>
<td class="r data">0.54345</td>
<td class="r data">0.02928</td>
<td class="r data">18.56</td>
<td class="r data">&lt;.0001</td>
<td class="r data">0.02936</td>
<td class="r data">18.51</td>
<td class="r data">&lt;.0001</td>
</tr>
<tr>
<th class="l rowheader" scope="row">rxlev</th>
<th class="r data">1</th>
<td class="r data" nowrap>-0.02907</td>
<td class="r data">0.04158</td>
<td class="r data" nowrap>-0.70</td>
<td class="r data">0.4846</td>
<td class="r data">0.04160</td>
<td class="r data" nowrap>-0.70</td>
<td class="r data">0.4847</td>
</tr>
<tr>
<th class="l rowheader" scope="row">rxlevplus</th>
<th class="r data">1</th>
<td class="r data" nowrap>-0.13176</td>
<td class="r data">0.04179</td>
<td class="r data" nowrap>-3.15</td>
<td class="r data">0.0017</td>
<td class="r data">0.04172</td>
<td class="r data" nowrap>-3.16</td>
<td class="r data">0.0016</td>
</tr>
</tbody>
</table>
</div>
